Calculating life expectancy and exploring missingness

### 2017-01-20 Code/1_Explore_Data.Rmd

```{r load_libs, echo=F, warning=F, message=F}
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

@@

```{r read_data}
dat <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/deaths-cause-6913.txt", header = F,sep="\t")
names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
```

```{r recode_variables}
dat <- dat %>%
  mutate(State2 = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         Age2 = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         Age3 = 0*(Age2 == "<1 year") + 1*(Age2 == "1-4 years") + 5*(Age2 == "5-9 years") +
                10*(Age2 == "10-14 years") + 15*(Age2 == "15-19 years") + 20*(Age2 == "20-24 years") + 
                25*(Age2 == "25-29 years") + 30*(Age2 == "30-34 years") + 35*(Age2 == "35-39 years") +
                40*(Age2 == "40-44 years") + 45*(Age2 == "45-49 years") + 50*(Age2 == "50-54 years") + 
                55*(Age2 == "55-59 years") + 60*(Age2 == "60-64 years") + 65*(Age2 == "65-69 years") +   
                70*(Age2 == "70-74 years") + 75*(Age2 == "75-79 years") + 80*(Age2 == "80-84 years") + 
                85*(Age2 == "85+ years"),
         nx = 1*(Age2 == "<1 year") + 4*(Age2 == "1-4 years") + 5*(Age2 != "<1 year" & Age2 != "1-4 years"),
         Sex2 = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
         Race2 = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
         COD2 = factor(COD, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
                                                        "Non-communicable", "Injuries", "All other causes")), 
         Year2 = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
         Year3 = as.numeric(as.character(Year2)),
         RaceSex = interaction(Race2, Sex2, sep = " "),
         StateYearSex = interaction(State2, Year2, Sex2),
         StateYearRaceSex = interaction(State2, Year2, Race2, Sex2))
```

```{r unknown_age_group}
#there is an unknown age group for each combo of state-year-race-sex-age:
table(dat$Age2)

#Count of death is always equal to 0 for unknown age groups
table(dat$Age, dat$Count==0)

#remove these data from the data we analyze:
dat.clean <- dat[dat$Age2 != "Unknown", ]
```

@@

```{r explore_dataset}
table(dat.clean$Population == 0, useNA = "always")
#654 cells have Populations of 0.

@@

summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
#whenever the population is 0, the Crude_Rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?

#States with at least one strata of 0
na.pops <- dat.clean %>% 
  select(State2, Population) %>% 
  filter(Population == 0)

unique(na.pops$State2)

pops.le.20 <- dat.clean %>% select(State2, Population) %>% filter(Population <= 20)
unique(pops.le.20$State2)
```

**Missingness of Count (of deaths)**
If there are between 1 and 9 deaths in the strata, this information is 'coarsened' and coded as NA. Later in this document we investigate trends in supression/missingness to understand how it differs according to state, age, race, and sex.

Since our goal is to compute the difference in life expectancy between Blacks and Whites by strata, we impute death counts at their logical extremes to explore this affects the results.

- Set all missing counts to min({1, 5, and 9}, population_size).

@@

```{r addvar_cod_proportions}
dat.clean <- dat.clean %>% group_by(State2, RaceSex, Year3, Age3) %>%
                           arrange(State2, RaceSex, Year3, Age3) %>%
                           mutate(total_deaths1 = sum(Count_md1),
                                  total_deaths5 = sum(Count_md5),
                                  total_deaths9 = sum(Count_md9),
                                  cod_prop_death1 = Count_md1/total_deaths1, 
                                  cod_prop_death5 = Count_md5/total_deaths5,
                                  cod_prop_death9 = Count_md9/total_deaths9)

@@

```{r aggregate_over_COD}
dat2 <- dat.clean %>% group_by(State2, Year2, RaceSex, Age2) %>%
  arrange(State2, RaceSex, Year2, Age2) %>%
  summarise(Population = first(Population), 
            Race2 = first(Race2), 
            Sex2 = first(Sex2), 
            StateYearRaceSex = first(StateYearRaceSex),
            nx = first(nx),
            Age3 = first(Age3),
            Year3 = first(Year3),
            total_deaths1 = first(total_deaths1),
            total_deaths5 = first(total_deaths5),
            total_deaths9 = first(total_deaths9),
            cod_prop_death1 = first(cod_prop_death1),
            cod_prop_death5 = first(cod_prop_death5),
            cod_prop_death9 = first(cod_prop_death9))

#return stratum ids with any age-specific population counts of 0.
#unique(dat2$StateYearRaceSex[dat2$Population == 0])
non.zero.stratums <- setdiff(levels(dat2$StateYearRaceSex), unique(dat2$StateYearRaceSex[dat2$Population == 0]))
```

Add the total population (across age brackets) to the dat2 dataset.

```{r create_var_pop_across_age}
dat3 <- dat2 %>% ungroup() %>% 
  group_by(StateYearRaceSex) %>%
  summarise(Pop_across_age = sum(Population))

dat2 <- merge(dat2, dat3, by = "StateYearRaceSex")
rm(dat3)
dat2 <- dat2 %>% arrange(State2, Year3, Race2, Sex2, Age3)
```

In those strata that do not have any age-specific population counts of 0, create a life table and store it in a list.

```{r calculate_life_expectancy}
#takes 10 mins for all three

list.life.tables <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
#list.life.tables1 <- list.life.tables5 <- list.life.tables9 <- vector(mode = "list", length = length(non.zero.stratums))
list.index <- 1

system.time(
for(stratum.id in non.zero.stratums) {
#for(stratum.id in head(non.zero.stratums)) {

  list.life.tables[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2),   
  #list.life.tables1[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
                                               num.ages.in.group = "nx", 
                                               death.counts = "total_deaths1", 
                                               population.counts = "Population")

  list.life.tables5[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
                                               num.ages.in.group = "nx", 
                                               death.counts = "total_deaths5", 
                                               population.counts = "Population")

  list.life.tables9[[list.index]] <- life.table(data = subset(dat2, StateYearRaceSex == stratum.id) %>% arrange(Age2), 
                                               num.ages.in.group = "nx", 
                                               death.counts = "total_deaths9", 
                                               population.counts = "Population")
  list.index <- list.index + 1
  print(list.index)
  }
)
```

Add the life expectancy at birth to the data file:

```{r add_le_to_dataset}
le <- le5 <- le9 <- rep(NA, length(non.zero.stratums))

for (i in 1:length(non.zero.stratums)) {
  le[i] <- list.life.tables[[i]]$e_x[1]
  le5[i] <- list.life.tables5[[i]]$e_x[1]
  le9[i] <- list.life.tables9[[i]]$e_x[1]
}

les <- data.frame(life.expectancy.birth = le, 
                  life.expectancy.birth.5 = le5, 
                  life.expectancy.birth.9 = le9, 
                  StateYearRaceSex = non.zero.stratums)

dat2 <- merge(dat2, les, by = "StateYearRaceSex", all.x =T)

rm(les)

ggplot(dat2 %>% filter(Age2 == "<1 year"), aes(x = life.expectancy.birth)) + 
  geom_density(aes(fill="Impute 1"), alpha = 0.3) + 
  geom_density(aes(x = life.expectancy.birth.5, fill = "Impute 5"), alpha = 0.3) +
  geom_density(aes(x = life.expectancy.birth.9, fill="Impute 9"), alpha = 0.3) +
  facet_grid(Sex2 ~ Race2)
```

Make a new dataset that contains information on the gap:

```{r create_dataset_BlackWhite}
#this dataset contains the black-white data within the same row (wide format) for each year-state-sex strata

Black <- dat2 %>% filter(Race2 == "Black" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, 
                         life.expectancy.birth, life.expectancy.birth.5, 
                         life.expectancy.birth.9) %>%
                  rename(Pop_black = Pop_across_age,  
                         le.birth.black = life.expectancy.birth, 
                         le.birth.black5 = life.expectancy.birth.5, 
                         le.birth.black9 = life.expectancy.birth.9)

White <- dat2 %>% filter(Race2 == "White" & Age3 == 0) %>% 
                  select(State2, Year3, Sex2, Pop_across_age, 
                         life.expectancy.birth, life.expectancy.birth.5, 
                         life.expectancy.birth.9) %>%
                  rename(Pop_white = Pop_across_age,  
                         le.birth.white = life.expectancy.birth,
                         le.birth.white5 = life.expectancy.birth.5,
                         le.birth.white9 = life.expectancy.birth.9)

BlackWhite <- merge(Black, White, by = c("State2", "Year3", "Sex2"))

BlackWhite <- BlackWhite %>% mutate(WBgap = le.birth.white - le.birth.black,
                                    WBgap5 = le.birth.white5 - le.birth.black5,
                                    WBgap9 = le.birth.white9 - le.birth.black9)

rm(Black, White)
```

Decompose the gap by age. That is, for each year-state-sex combination, compare the life table for blacks to whites and decompose the gap in life expectancy at birth by age.

```{r create_paired_ids_dataset}
#paired.ids will contain the index names that are sent to the age decomposition function in the next code chunk

#first, create list of stratum.ids for each race-sex combination
for (rs in unique(levels(dat.clean$RaceSex))) {
  assign(paste0(gsub(" ", "", rs), ".stratum.ids"), 
         dat.clean %>% 
           ungroup() %>%
           filter(RaceSex == rs) %>%
           select(StateYearRaceSex, StateYearSex, Sex2, State2, Year3) %>%
           group_by(StateYearRaceSex, StateYearSex) %>%
           filter(row_number() == 1)
         )

}

#then, merge the black-white lists
Female.ids <- merge(WhiteFemale.stratum.ids, BlackFemale.stratum.ids[, -c(3:5)], 
                    by = "StateYearSex", suffixes = c("White", "Black"))
head(Female.ids)

Male.ids <- merge(WhiteMale.stratum.ids, BlackMale.stratum.ids[, -c(3:5)],
                  by = "StateYearSex", suffixes = c("White", "Black"))
head(Male.ids)

paired.ids <- rbind(Male.ids, Female.ids)
```

```{r calculate_age_decomp}
#runs in about a minute

list.age.decomp.tables <- list.age.decomp.tables5 <- list.age.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])

for(i in 1:dim(paired.ids)[1]) {
  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])

  if(length(white.index) != 0 & length(black.index) != 0) {
    list.age.decomp.tables[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")

    list.age.decomp.tables5[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables5[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables5[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")

    list.age.decomp.tables9[[i]] <- le_age_decomp(life.table1 = as.data.frame(list.life.tables9[[black.index]]), 
                                                 name.lt1 = "Black", 
                                                 life.table2 = as.data.frame(list.life.tables9[[white.index]]), 
                                                 name.lt2 = "White", age.groups = "Age3")
  } else {
    list.age.decomp.tables[[i]] <- list.age.decomp.tables5[[i]] <- list.age.decomp.tables9[[i]] <-"Couldn't compare life tables"
  }
}

@@

```{r calculate_cod_decomp}
#stopped here - need to figure this out
#if the count of deaths == 0 for all causes of death within an age band then the COD proportion is NA
#this messes up the cod decomposition. What to do?
#example - no black people dying between the ages of 5-9 of anything. but have whites dying in that age group so their
#COD proportions are okay.

#takes 1.5 hours for all three
list.cod.decomp.tables <- list.cod.decomp.tables5 <- list.cod.decomp.tables9 <- vector(mode = "list", length = dim(paired.ids)[1])

system.time(
for(i in 1:dim(paired.ids)[1]) {
  white.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexWhite"])
  black.index <- which(non.zero.stratums == paired.ids[i, "StateYearRaceSexBlack"])

  if(length(white.index) != 0 & length(black.index) != 0) {
    ct.white <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexWhite"]) %>% 
                ungroup() %>%
                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
                rename(prop.white1 = cod_prop_death1, prop.white5 = cod_prop_death5, prop.white9 = cod_prop_death9)

    ct.black <- subset(dat.clean, StateYearRaceSex == paired.ids[i, "StateYearRaceSexBlack"]) %>% 
                ungroup() %>%
                select(Age2, COD2, cod_prop_death1, cod_prop_death5, cod_prop_death9) %>%
                rename(prop.black1 = cod_prop_death1, prop.black5 = cod_prop_death5, prop.black9 = cod_prop_death9)
    ct.both <- merge(ct.white, ct.black, by = c("Age2", "COD2")) 
    ct.both <- ct.both %>% arrange(COD2, Age2)

    if(any(is.na(ct.both$prop.black1)) == F & any(is.na(ct.both$prop.white1)) == F) { #ensures that there are no ages with 0 deaths - for these COD prop == 0 and can't compute

        list.cod.decomp.tables[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white1, prop.black1), 
                                                              age.colname.cod.table = "Age2", 
                                                              COD.colname.cod.table = "COD2", 
                                                              prop1.colname.cod.table = "prop.black1", 
                                                              prop2.colname.cod.table = "prop.white1")

        list.cod.decomp.tables5[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white5, prop.black5), 
                                                              age.colname.cod.table = "Age2", 
                                                              COD.colname.cod.table = "COD2", 
                                                              prop1.colname.cod.table = "prop.black5", 
                                                              prop2.colname.cod.table = "prop.white5")

        list.cod.decomp.tables9[[i]] <- cause_of_death_decomp(life.table1 = as.data.frame(list.life.tables[[black.index]]),
                                                              life.table2 = as.data.frame(list.life.tables[[white.index]]), 
                                                              decomp.table = as.data.frame(list.age.decomp.tables[[i]]),
                                                              cod.table = ct.both %>% select(Age2, COD2, prop.white9, prop.black9), 
                                                              age.colname.cod.table = "Age2", 
                                                              COD.colname.cod.table = "COD2", 
                                                              prop1.colname.cod.table = "prop.black9", 
                                                              prop2.colname.cod.table = "prop.white9")

    } else { list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Encountered age group with no deaths"

    }
  } else {

    list.cod.decomp.tables[[i]] <- list.cod.decomp.tables5[[i]] <- list.cod.decomp.tables9[[i]] <- "Couldn't compare life tables"
  }

  print(i)
}
)
```

```{r  update_decomp_for_plotting}
#update the decomposition table for plotting purposes
for(i in 1:dim(paired.ids)[1]) {
    if(list.cod.decomp.tables[[i]][1] != "Couldn't compare life tables" & 
       list.cod.decomp.tables[[i]][1] != "Encountered age group with no deaths") {

       list.cod.decomp.tables[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death", 
                                                            sign.var = "sign", decomp.var = "C_xi", 
                                                            decomp.var.prop = "C_xi_proportion")

       list.cod.decomp.tables5[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables5[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death", 
                                                            sign.var = "sign", decomp.var = "C_xi", 
                                                            decomp.var.prop = "C_xi_proportion")

       list.cod.decomp.tables9[[i]] <- make_dataset_cod_plot(cod.decomp.table = list.cod.decomp.tables9[[i]],
                                                            age.groups = "Ages", 
                                                            cause.of.death = "Cause.of.death", 
                                                            sign.var = "sign", decomp.var = "C_xi", 
                                                            decomp.var.prop = "C_xi_proportion") 
    } 
  print(i)
}  
```

```{r pretty_cod}
ggplotly(ggplot(list.cod.decomp.tables[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))

ggplotly(ggplot(list.cod.decomp.tables9[[1]], aes(y = Ages, x = start)) + 
           geom_segment(aes(xend = finish, col = Cause.of.death, yend = Ages), lwd = 4) +
           xlab("Contribution to life expectancy gap") +
           ylab("Age group\n") + theme_minimal() +
           geom_vline(xintercept = 0))
```

###Initial graphs on missing data
Notes on Figure 1 (below):

- Figure 1 is sorted by size of the black population in the state in 2013. This is so you can see how missingness is a function of population size. We expect that states with less black people will have more missing data and more counts of exactly zero. Thus, states with the biggest populations (of blacks and whites) will "hold constant" the effect of population size on the amount of missing data.

- The first panels you see in Figure 1 (below) are for the states with the smallest black populations. For whites, the amount of missing data increases in age until about 40 years old, and then decreases. For blacks, the amount of missing data is very low across age. This seemed strange to me at first, but then I remembered that the data is not missing if the count is equal to 0, which will be the case for many strata in these states with the lowest black population.

- As you read left to right (as the black population size increases by state), the trends in missing data across age look more similar by race, although blacks appear to have more missing data in individuals < 30 years old for the states with the most black people.

**Figure 1: Proportion missing by State (panel) and Age group (x-axis), ordered by increasing size of black population in 2013**
```{r bystate, fig.width = 12, fig.height = 37, echo=F}
bystate <- 
  dat %>%
  group_by(State2, Race2, Age2) %>%
  summarise(prop.missing = mean(is.na(Count)))

#recode a new count variable that "counts the missing data" so that these rows are used when calculating the % of data == 0.
dat$Count2 <- dat$Count
dat$Count2[is.na(dat$Count) == T] <- 1

bystate2 <- 
  dat %>%
  group_by(State2, Race2, Age2) %>%
  summarise(prop.zero = mean(Count2 == 0))

bystate <- merge(bystate, bystate2, by=c("State2", "Race2", "Age2"))

#calculate the size of the black population in 2013 by State
bystate.bpop <- 
  dat %>%
  filter(COD2 == "Cancers" & Year2 == 2013 & Race2 == "Black") %>% #to stop repeats of the population size
  group_by(State2) %>%
  summarize(pop_by_race = sum(as.numeric(Population)))

#calculate the size of the black population in 1969 by State
bystate.bpop.1969 <- 
  dat %>%
  filter(COD2 == "Cancers" & Year2 == 1969 & Race2 == "Black") %>% #to stop repeats of the population size
  group_by(State2) %>%
  summarize(pop_by_race_1969 = sum(as.numeric(Population)))

bystate <- merge(bystate, bystate.bpop, by = "State2", all.x = T)
bystate <- merge(bystate, bystate.bpop.1969, by = "State2", all.x = T)
bystate$State3 <- reorder(bystate$State2, bystate$pop_by_race)

ggplot(bystate, aes(x = Age2, y = prop.missing)) + 
  geom_point(aes(col=Race2)) + facet_wrap(~ State3, ncol = 4) +
  theme_bw(base_size = 11) +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing (%)") +
  xlab("Age group")

ggplot(subset(bystate, Age2 == "<1 year"), aes(pop_by_race_1969, pop_by_race)) + geom_point() + geom_abline(intercept = 0, slope = 1)

unique(bystate$State2[bystate$Age2 == "<1 year" & bystate$pop_by_race_1969 < 10000])
```

Notes on Figure 2 (below):

- Figure 2 examines the proportion of counts equal to zero. The purpose of this examination is to visually confirm that there are higher proportions of zero counts in the states with the least amount of blacks. This is very clear when viewing the figure.

- Question - any explantion for why the portion of zeroes stays relatively higher in blacks vs. whites in the states with the most blacks (within the younger age groups)?

**Figure 2: Proportion zero by State (panel) and Age group (x-axis), ordered by increasing size of black population in 2013**

```{r zero_by_state, fig.width = 12, fig.height = 37, echo=F}
ggplot(bystate, aes(x = Age2, y = prop.zero)) + 
  geom_point(aes(col=Race2)) + facet_wrap(~ State3, ncol = 4) +
  theme_bw(base_size = 11) +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Zero (%)") +
  xlab("Age group") 
```

###Other plots (less interesting/informative)

**How is missingness related to age, race, and sex?**

```{r missing_data, fig.width = 10, fig.height = 5, echo=F}
#table(is.na(dat$Count))

table1.whitemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White Male"]), 
                                           dat$Age2[dat$RaceSex == "White Male"]),2)*100,2)
table1.blackmale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black Male"]), 
                                           dat$Age2[dat$RaceSex == "Black Male"]),2)*100,2)
table1.whitefemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "White Female"]), 
                                           dat$Age2[dat$RaceSex == "White Female"]),2)*100,2)
table1.blackfemale <- round(prop.table(table(is.na(dat$Count[dat$RaceSex == "Black Female"]), 
                                           dat$Age2[dat$RaceSex == "Black Female"]),2)*100,2)

miss1 <- data.frame("white.male" = table1.whitemale[2, ], 
                    "black.male"= table1.blackmale[2, ],
                    "white.female" = table1.whitefemale[2, ], 
                    "black.female"= table1.blackfemale[2, ],
                    "age" = names(table1.whitemale[2, ]))
miss1$age <- factor(miss1$age, levels(miss1$age)[c(1, 2, 11, 3:10, 12:20)])

miss1 <- gather(miss1, racesex, missingness, c(white.male, black.male, white.female, black.female))

ggplot(miss1, aes(x = age, y = missingness)) + geom_point() + geom_point(aes(col=racesex)) +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing (%)") +
  xlab("Age group")
```

**Let's look at the proportion with 0's**

```{r zero_data, fig.width = 10, fig.height = 5, echo=F}
table2.whitemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White Male"]==0, 
                                       dat$Age2[dat$RaceSex == "White Male"]),2)*100,2)
table2.blackmale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black Male"]==0, 
                                       dat$Age2[dat$RaceSex == "Black Male"]),2)*100,2)
table2.whitefemale <- round(prop.table(table(dat$Count[dat$RaceSex == "White Female"]==0, 
                                       dat$Age2[dat$RaceSex == "White Female"]),2)*100,2)
table2.blackfemale <- round(prop.table(table(dat$Count[dat$RaceSex == "Black Female"]==0, 
                                       dat$Age2[dat$RaceSex == "Black Female"]),2)*100,2)

miss2 <- data.frame("whitemale" = table2.whitemale[2, ], 
                    "blackmale"= table2.blackmale[2, ],
                    "whitefemale" = table2.whitefemale[2, ], 
                    "blackfemale"= table2.blackfemale[2, ],
                    "age" = names(table2.blackmale[2, ]))

miss2$age <- factor(miss2$age, levels(miss2$age)[c(1, 2, 11, 3:10, 12:20)])

>>>>

```{r load_libraries, echo=F, warning=F, message=F}
library(tidyverse)
```

@@

```{r read_raw_data}
dat <- read.table("/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/deaths-cause-6913.txt", header = F,sep="\t")
names(dat) <- c("State", "Age", "Sex", "Race", "COD", "Year", "Crude_Rate", "Count", "Population")
# names and order of variables in the dataset are noted in the .dic dictionary file found in the same dir as the dataset.
```

```{r recode_variables}
dat <- dat %>%
  mutate(State_factor = factor(State, levels = c(0:50), labels = c("Alabama", "Alaska", "Arizona", "Arkansas","California",
                                                             "Colorado", "Connecticut", "Delaware", "Washington DC",
                                                             "Florida", "Georgia", "Hawaii","Idaho", "Illinois", 
                                                             "Indiana","Iowa", "Kansas", "Kentucky", "Louisiana", 
                                                             "Maine", "Maryland", "Massachusetts", "Michigan", 
                                                             "Minnesota", "Mississippi","Missouri", "Montana", 
                                                             "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                                                             "New Mexico", "New York","North Carolina","North Dakota",
                                                             "Ohio", "Oklahoma","Oregon", "Pennsylvania", 
                                                             "Rhode Island", "South Carolina", "South Dakota", 
                                                             "Tennessee", "Texas", "Utah", "Vermont", "Virginia",
                                                             "Washington", "West Virginia", "Wisconsin", "Wyoming")), 
         Age_factor = factor(Age, levels = c(0:19), labels = c("<1 year", "1-4 years", "5-9 years", "10-14 years", 
                                                         "15-19 years", "20-24 years", "25-29 years", "30-34 years",
                                                         "35-39 years", "40-44 years", "45-49 years", "50-54 years", 
                                                         "55-59 years", "60-64 years", "65-69 years", "70-74 years", 
                                                         "75-79 years", "80-84 years","85+ years", "Unknown"), ordered = T),
         Age_minbin = 0*(Age_factor == "<1 year") + 1*(Age_factor == "1-4 years") + 5*(Age_factor == "5-9 years") +
                10*(Age_factor == "10-14 years") + 15*(Age_factor == "15-19 years") + 20*(Age_factor == "20-24 years") + 
                25*(Age_factor == "25-29 years") + 30*(Age_factor == "30-34 years") + 35*(Age_factor == "35-39 years") +
                40*(Age_factor == "40-44 years") + 45*(Age_factor == "45-49 years") + 50*(Age_factor == "50-54 years") + 
                55*(Age_factor == "55-59 years") + 60*(Age_factor == "60-64 years") + 65*(Age_factor == "65-69 years") +   
                70*(Age_factor == "70-74 years") + 75*(Age_factor == "75-79 years") + 80*(Age_factor == "80-84 years") + 
                85*(Age_factor == "85+ years"),
         Age_bin_size = 1*(Age_factor == "<1 year") + 4*(Age_factor == "1-4 years") + 5*(Age_factor != "<1 year" & Age_factor != "1-4 years"),
         Sex_factor = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
         Race_factor = factor(Race, levels = c(0, 1), labels = c("White", "Black")),
         COD_factor = factor(COD, levels = c(0:5), labels = c("Cardiovascular", "Cancers", "Communicable", 
                                                        "Non-communicable", "Injuries", "All other causes")), 
         Year_factor = factor(Year, levels = c(0:44), labels = as.character(1969:2013)),
         Year_actual = as.numeric(as.character(Year_factor)),
         RaceSex = interaction(Race_factor, Sex_factor, sep = " "),
         StateYearSex = interaction(State_factor, Year_factor, Sex_factor),
         StateYearRaceSex = interaction(State_factor, Year_factor, Race_factor, Sex_factor))

head(dat)
```

```{r remove_unknown_age_group}
#there is an unknown age group for each combo of state-year-race-sex-age:
table(dat$Age_factor)

#Count of death is always equal to 0 for unknown age groups
table(dat$Age_factor, dat$Count==0)

#remove these data from the data we analyze:
dat.clean <- dat[dat$Age_factor != "Unknown", ]
```

@@

```{r recode_pops_of_zero}
table(dat.clean$Population == 0, useNA = "always")
#654 cells have Populations of 0.

@@

summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
#whenever the population is 0, the Crude_Rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?
```

```{r list_states_0_le20_popcounts}
#States with at least one strata of 0
na.pops <- dat.clean %>% 
  select(State_factor, Population) %>% 
  filter(Population == 0)

unique(na.pops$State_factor)

pops.le.20 <- dat.clean %>% select(State_factor, Population) %>% filter(Population <= 20)
unique(pops.le.20$State_factor)
```

**Missingness of Count (of deaths)**
If there are between 1 and 9 deaths in the strata, this information is 'coarsened' and the number of deaths (Count) is coded as NA. Later in this document we investigate trends in supression/missingness to understand how it differs according to state, age, race, and sex.

Since our goal is to compute the difference in life expectancy between Blacks and Whites by strata, we impute death counts at their logical extremes to explore how this affects the results. Further, we use this rough data to create the shiny app while the Bayesian model is being developed/run.

@@

```{r addvar_cod_proportions} 
# Add the proportions of deaths due to each cause as a new variable.
# This variable is used to decompose the life expectancy gap by cause of death.

dat.clean <- dat.clean %>% group_by(State_factor, RaceSex, Year_actual, Age_minbin) %>%
                           arrange(State_factor, RaceSex, Year_actual, Age_minbin) %>%
                           mutate(total_deaths1 = sum(Count_md1),
                                  total_deaths5 = sum(Count_md5),
                                  total_deaths9 = sum(Count_md9),
                                  cod_prop_death1 = Count_md1/total_deaths1, 
                                  cod_prop_death5 = Count_md5/total_deaths5,
                                  cod_prop_death9 = Count_md9/total_deaths9)

@@

```{r aggregate_over_COD}
dat.aggregated <- dat.clean %>% group_by(State_factor, Year_factor, RaceSex, Age_factor) %>%
                                arrange(State_factor, RaceSex, Year_factor, Age_factor) %>%
                                summarise(Population = first(Population), 
                                          Race_factor = first(Race_factor), 
                                          Sex_factor = first(Sex_factor), 
                                          StateYearRaceSex = first(StateYearRaceSex),
                                          Age_bin_size = first(Age_bin_size),
                                          Age_minbin = first(Age_minbin),
                                          Year_actual = first(Year_actual),
                                          total_deaths1 = first(total_deaths1),
                                          total_deaths5 = first(total_deaths5),
                                          total_deaths9 = first(total_deaths9),
                                          cod_prop_death1 = first(cod_prop_death1),
                                          cod_prop_death5 = first(cod_prop_death5),
                                          cod_prop_death9 = first(cod_prop_death9))

#return "stratum ids" with any age-specific population counts of 0.
non.zero.stratums <- setdiff(levels(dat.aggregated$StateYearRaceSex), 
                             unique(dat.aggregated$StateYearRaceSex[dat.aggregated$Population == 0]))
```

Add the total population (across age brackets) to the dat.aggregated dataset.

```{r create_var_pop_across_age}
dat3 <- dat.aggregated %>% ungroup() %>% 
                           group_by(StateYearRaceSex) %>%
                           summarise(Pop_across_age = sum(Population))

dat.aggregated <- merge(dat.aggregated, dat3, by = "StateYearRaceSex")

rm(dat3)

dat.aggregated <- dat.aggregated %>% arrange(State_factor, Year_actual, Race_factor, Sex_factor, Age_minbin)
```

```{r write_data_to_csvs}
#save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
write_csv(dat.clean, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_clean.Rdata", col_names = T)
write_csv(dat.aggregated, path = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/dat_aggregated.Rdata", col_names = T)

@@

```{r save_specified_files}
# save(dat.clean, dat2, BlackWhite, BlackWhiteSmooth, paired.ids, list.age.decomp.tables, list.cod.decomp.tables,
#      list.age.decomp.tables5, list.cod.decomp.tables5, list.age.decomp.tables9, list.cod.decomp.tables9,
#      list.age.decomp.tables.smoothed, list.cod.decomp.tables.smoothed, list.cod.marginal.tables.smoothed,
#      non.zero.stratums, non.zero.stratums.smoothed, cod.contributions,
#      file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/main_datasets.Rdata")
```
